---
title: "Metabolic Interaction Network Analysis: ENBI, Cross-feeding, and Engraftment"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5, dpi = 150)
```

# Setup

```{r libraries}
library(tidyverse)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggsignif)
library(Hmisc)   # rcorr for correlation p-values
library(broom)

setwd("/Users/guillaumeurtecho/Dropbox/UrtechoLab/Projects/2024/FMT_Metabolic_models_ZZ/Github_Workshop/Figure4D")

```

```{r file-paths}
# ── FILE PATHS ────────────────────────────────────────────────────────────────
# UPDATE interactions_file to point to the full MICOM interactions output.
# The file should have columns: sample_id, focal, partner, class, flux,
# mass_flux, C_flux, N_flux, n_ints (same schema as ExcerptInteractions).
interactions_file   <- "input/interactions_western_openbiome_summary.csv"       
taxonomy_file       <- "../utils/full_taxonomy.csv"
engraftment_file    <- "../Figure3CE/data/MedianEngraftmentScores_Openbiome.csv"
genome_scores_file  <- "../OpenBiome_Predictions/input/model_files/genome_scores.tsv"
superdonor_file     <- "../OpenBiome_Predictions/output/SuperdonorEngraftmentScores.csv"
```

```{r highlighted-samples}
# Highlighted donor microbiomes
highlight_ids <- c(
  superdonor_ob = "SRR9224383",
  superdonor2   = "SRR9224196",
  worst1        = "SRR9224548",
  secondworst   = "SRR9224546"
)

highlight_labels <- c(
  SRR9224383 = "Super-donor",
  SRR9224196 = "Super-donor 2",
  SRR9224548 = "Poor Donor 1",
  SRR9224546 = "Poor Donor 2"
)

highlight_colors <- c(
  SRR9224383 = "firebrick2",
  SRR9224196 = "salmon",
  SRR9224548 = "steelblue3",
  SRR9224546 = "lightblue3"
)


```

```{r load-data}
interactions <- read_csv(interactions_file, show_col_types = FALSE)
taxonomy     <- read_csv(taxonomy_file, show_col_types = FALSE)
engraftment  <- read_csv(engraftment_file, show_col_types = FALSE)
genome_scores <- read_tsv(genome_scores_file, show_col_types = FALSE)
superdonor   <- read_csv(superdonor_file, show_col_types = FALSE)

cat("Interactions:", nrow(interactions), "rows,",
    n_distinct(interactions$sample_id), "samples\n")
cat("Engraftment scores:", nrow(engraftment), "donors\n")
```


# Part 1: Community-level Interaction Balance vs Engraftment

The net interaction balance (rho) follows Corral Lopez et al. (2026, *Science*):

$$\rho = \frac{\sum \text{cross-feeding flux} - \sum \text{competitive flux}}{\sum \text{cross-feeding flux} + \sum \text{competitive flux}}$$

where cross-feeding = `provided` + `received` and competition = `co-consumed`.
Because we lack a canonical healthy baseline, we report raw rho per sample and
also compute an ENBI-like score using the cohort median rho as reference.

```{r part1-calculate-rho}
# Classify interaction types and aggregate per sample
sample_metrics <- interactions %>%
  mutate(
    int_type = case_when(
      class == "co-consumed" ~ "competition",
      class %in% c("provided", "received") ~ "cross_feeding"
    )
  ) %>% #Label as competition or cross-feeding
  group_by(sample_id, int_type) %>%
  summarise(total_flux = sum(flux, na.rm = TRUE), .groups = "drop") %>% #sum flux of competition vs crossfeeding
  pivot_wider(names_from = int_type, values_from = total_flux,
              values_fill = 0) %>%
  mutate(
    rho  = (cross_feeding - competition) / (cross_feeding + competition), #calculate net interaction balance
    enbi = rho - median(rho, na.rm = TRUE) #center around median
  )

cat("Samples with rho calculated:", nrow(sample_metrics), "\n")
summary(sample_metrics$rho)
```

```{r part1-merge-engraftment}
# The engraftment file uses "donor" as the ID column.
# Most entries are SRR IDs that match sample_id directly.
sample_eng <- sample_metrics %>%
  inner_join(engraftment, by = c("sample_id" = "donor")) %>%
  mutate(
    highlight = if_else(sample_id %in% highlight_ids, sample_id, NA_character_),  #label samples
    label     = highlight_labels[highlight]
  )
```



ENBI vs engraftment

```{r}
cor_rho <- cor.test(sample_eng$rho, sample_eng$med_engraft,
                    method = "spearman")

# Format p-value: replace machine-zero with < 2.2e-16
p_fmt <- if (cor_rho$p.value == 0) {
  "italic(p) < 2.2 %*% 10^{-16}"
} else {
  paste0("italic(p) == ", signif(cor_rho$p.value, 3))
}

rho_label <- paste0("rho == ", round(cor_rho$estimate, 3))

p1a <- ggplot(sample_eng, aes(x = rho, y = med_engraft)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey50",
              linewidth = 0.6, linetype = "dashed") +
  geom_point(aes(color = highlight), size = 2.2, alpha = 0.8) +
  geom_point(data = filter(sample_eng, !is.na(highlight)),
             aes(color = highlight), size = 3.5) +
  geom_text_repel(data = filter(sample_eng, !is.na(highlight)),
                  aes(label = label, color = highlight),
                  size = 4, fontface = "bold",
                  box.padding = 0.5, max.overlaps = 20) +
  scale_color_manual(values = highlight_colors, na.value = "grey60",
                     guide = "none") +
  annotate("text", x = -Inf, y = Inf, label = rho_label,
           parse = TRUE, hjust = -0.05, vjust = 1.3, size = 5) +
  annotate("text", x = -Inf, y = Inf, label = p_fmt,
           parse = TRUE, hjust = -0.05, vjust = 3.0, size = 5) +
  labs(x = "Ecological Network Balance Index (ENBI)",
       y = "Median engraftment score") +
  theme_cowplot()

p1a

ggsave("figs/Fig4D_ENBIvsEngraftment.pdf", width = 3, height = 3)

```

