---
title: "figure_2"
output: html_document
date: "2025-08-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dir.create("../figures", showWarnings = FALSE, recursive = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(cowplot)
library(ggrepel)
library(patchwork)
library(ggpubr)
library(phyloseq)
library(DESeq2)

pf <- read.csv('../../utils/colorPalette.csv', stringsAsFactors = FALSE)
palette_vec <- setNames(trimws(pf$x), trimws(pf$X))
```


```{r figure 2A}
raw <- read.csv('../data/abundance_raw_subset_filtered.csv', check.names = FALSE)
taxonomy <- read.csv('../../utils/full_taxonomy.csv', check.names = FALSE)
DonorRecipientMapping <- read.csv('../../utils/DonorRecipientMapping.csv', check.names = FALSE)

sample_info_t6 <- DonorRecipientMapping %>% 
  select(recipient, clinical_response) %>%
  mutate(
    recipient = str_split_fixed(recipient, "_", 2)[,1]
  ) %>% 
  pivot_longer(
    cols      = c(recipient),
    names_to  = "type",
    values_to = "sample"        
  ) %>% mutate(
    sample = str_replace(sample, "-0", "-6"),
    time = "T6",
  ) %>% 
  select(-type) %>%
  column_to_rownames(var = "sample")

sample_info_t0 <- DonorRecipientMapping %>% 
  select(recipient, clinical_response) %>%
  mutate(
    recipient = str_split_fixed(recipient, "_", 2)[,1]
  ) %>%
  pivot_longer(
    cols      = c(recipient),
    names_to  = "type",
    values_to = "sample"        
  ) %>% 
  mutate(
    time = "T0"
  ) %>% 
  select(-type) %>%
  column_to_rownames(var = "sample")

sample_info <- bind_rows(sample_info_t0, sample_info_t6) %>% mutate(
  subject_id = str_split_fixed(rownames(.), "-", 2)[,1]
)

sample_info_R <- sample_info %>% filter(clinical_response == 'responder')
sample_info_NR <- sample_info %>% filter(clinical_response == 'non-responder')
rownames(taxonomy) <- taxonomy$taxon_id
taxonomy <- as.matrix(taxonomy)
rownames(raw) <- raw$taxon_id
raw <- raw %>% select(-taxon_id)
raw <- as.matrix(raw)

raw <- raw[, colnames(raw) %in% rownames(sample_info)]
sample_info <- sample_info[colnames(raw), , drop = FALSE]
OTU <- otu_table(raw, taxa_are_rows = TRUE)
TAX <- tax_table(taxonomy)
SAM <- sample_data(sample_info)
physeq <- phyloseq(OTU, TAX, SAM)

sample_data(physeq)$time <- factor(sample_data(physeq)$time, levels = c("T0", "T6"))
sample_data(physeq)$clinical_response <- factor(sample_data(physeq)$clinical_response, 
                                               levels = c("non-responder", "responder"))

sample_data(physeq)$subject_id <- factor(sample_data(physeq)$subject_id)
dds_resp <- phyloseq_to_deseq2(
  subset_samples(physeq, clinical_response == "responder"),
  ~ subject_id + time
) %>% DESeq()

dds_non_resp <- phyloseq_to_deseq2(
  subset_samples(physeq, clinical_response == "non-responder"),
  ~ subject_id + time
) %>% DESeq()

res_resp <- results(
  dds_resp,
  contrast = c("time", "T6", "T0"),
  alpha = 0.05,
  independentFiltering = FALSE,
  cooksCutoff = FALSE
) %>% as.data.frame() %>%
  rownames_to_column("taxon") %>%
  mutate(
    neglog10qval = -log10(padj),
    significance = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "enriched",
      padj < 0.05 & log2FoldChange < 0 ~ "depleted",
      TRUE ~ "n.s."
    )
  )

res_non_resp <- results(
  dds_non_resp,
  contrast = c("time", "T6", "T0"),
  alpha = 0.05,
  independentFiltering = FALSE,
  cooksCutoff = FALSE
) %>% as.data.frame() %>%
  rownames_to_column("taxon") %>%
  mutate(
    neglog10qval = -log10(padj),
    significance = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "enriched",
      padj < 0.05 & log2FoldChange < 0 ~ "depleted",
      TRUE ~ "n.s."
    )
  )

taxonomy_copy <- taxonomy %>%
  as.data.frame() %>%
  rownames_to_column('taxon')
res_resp <- res_resp %>%
  left_join(taxonomy_copy, by = 'taxon') 
  
res_non_resp <- res_non_resp %>%
  left_join(taxonomy_copy, by = 'taxon')

all_orders <- unique(c(
  res_resp$order[res_resp$significance %in% c("enriched", "depleted")],
  res_non_resp$order[res_non_resp$significance %in% c("enriched", "depleted")]
))

n_orders <- length(all_orders)
if (n_orders > 0) {
  if (n_orders <= 26) {
    order_colors <- setNames(
      palette.colors(n = n_orders, palette = "alphabet"),
      all_orders
    )
  } else {
    order_colors <- setNames(rainbow(n_orders), all_orders)
  }
} else {
  order_colors <- c()
}

order_colors <- c(order_colors, "n.s." = "gray")
res_resp <- res_resp %>%
  mutate(
    color_group = ifelse(significance %in% c("enriched", "depleted"), order, "n.s.")
  )

res_non_resp <- res_non_resp %>%
  mutate(
    color_group = ifelse(significance %in% c("enriched", "depleted"), order, "n.s.")
  )

all_levels <- c(all_orders, "n.s.")
res_resp$color_group <- factor(res_resp$color_group, levels = all_levels)
res_non_resp$color_group <- factor(res_non_resp$color_group, levels = all_levels)

shared_xlim <- c(-4, 6)
shared_ylim <- c(0, max(
  c(res_resp$neglog10qval, res_non_resp$neglog10qval), 
  na.rm = TRUE
))

pl1 <- ggplot(res_resp, aes(x = log2FoldChange, y = neglog10qval, color = color_group)) +
  geom_point(alpha = 0.7) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey80") +
  scale_color_manual(values = order_colors, name = "Order") +
  theme_cowplot() +
  xlab("log2(Normalized Effect)") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Responders") +
  annotate("text",
           x = shared_xlim[1], y = 2.5,
           label = "T0", hjust = 0, size = 6) +
  annotate("text",
           x = shared_xlim[2], y = 2.5,
           label = "T6", hjust = 1, size = 6) +  scale_x_continuous(breaks = seq(-15, 10, by = 2))

pl2 <- ggplot(res_non_resp, aes(x = log2FoldChange, y = neglog10qval, color = color_group)) +
  geom_point(alpha = 0.7) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey80") +
  scale_color_manual(values = order_colors, name = "Order") +
  theme_cowplot() +
  xlab("log2(Normalized Effect)") +
  ylab("-log10(adjusted p-value)") +
   ggtitle("Non-Responders") +
  annotate("text",
           x = shared_xlim[1], y = 2.5,
           label = "T0", hjust = 0, size = 6) +
  annotate("text",
           x = shared_xlim[2], y = 2.5,
           label = "T6", hjust = 1, size = 6) +
  scale_x_continuous(breaks = seq(-15, 10, by = 2))



pl1 <- pl1 + coord_cartesian(xlim = shared_xlim, ylim = shared_ylim)
pl2 <- pl2 + coord_cartesian(xlim = shared_xlim, ylim = shared_ylim)

fig_2a <- (pl1 + pl2) +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(title = "") &
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.5, "lines"),
                axis.text = element_text(size = 10))

if (n_orders > 10) {
  fig_2a <- fig_2a & guides(color = guide_legend(nrow = 3))
}

print(fig_2a)

ggsave(
  "../figures/2a.pdf",
  fig_2a,
  width = 5,
  height = 3.5,
  dpi = 600,
  bg = 'white'
)

```


```{r figure 2B}
total_t0_production = read.csv('../data/2A.csv', check.names = FALSE)

metabolites_to_plot <- c(
  'Valeric Acid' = 'M03134[e]',
  'Sulfite' = 'so3[e]',
  'Hydrogen' = 'h2[e]',
  'propionate' = 'ppa[e]',
  'acetate' = 'ac[e]',
  'butyrate' = 'but[e]',
  'thiosulfate(2-)' = 'tsul[e]',
  'Formate' = 'for[e]',
  'Methanethiol' = 'ch4s[e]'
)

palette <- c(
  'enriched' = 'red',
  'depleted' = '#008CFF',
  'n.s.' = 'gray'
)

fig_2b <- ggplot(total_t0_production, aes(x = log_fold_change, y = `-log10q_val`, color = enrichment)) + 
  geom_point(size = 1.5) + 
  geom_hline(yintercept = -log10(0.05), color = 'gray', linetype = 'dashed') + 
  geom_vline(xintercept = c(-1, 1), color = 'gray', linetype = 'dashed') +
  geom_text_repel(
  data = total_t0_production[total_t0_production$name %in% names(metabolites_to_plot), ],
  aes(label = name),
  size = 4,  # 2 Equivalent to fontsize=6 in matplotlib
  point.padding = 0.5,
  box.padding = 1.5,
  force = 1,
  arrow = arrow(length = unit(0.01, "npc"), type = "closed"),
  color = "black") + 
  scale_color_manual(values = palette) +
  scale_x_continuous(breaks = c(-5,-2.5,0,2.5,5,10,15)) +
  labs(
 # title = 'Differential Production Rates (T0) of IBS Patient vs. Donor',
  x = 'log2(Fold-Change)',
  y = '-log10(q-value)',
  color = 'Enrichment') + 
  theme_cowplot() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center title
    panel.grid.minor = element_blank(),
            axis.text = element_text(size = 10),
      axis.title = element_text(size = 12)
  ) +
  theme(plot.margin = margin(10, 10, 10, 10)) +
  theme_cowplot()

print(fig_2b)
ggsave(filename = '../figures/2b.pdf',
       plot = fig_2b,
       width = 5, 
       height = 3,
       dpi = 600,
       bg = 'white'
       )
```

```{r figure 2C}
metabolites_to_plot <- c(
  'Formate' = 'for[e]', 
  'acetate' = 'ac[e]', 
  'propionate' = 'ppa[e]', 
  'butyrate' = 'but[e]', 
 # 'Valeric Acid' = 'M03134[e]', 
  'Sulfite' = 'so3[e]', 
  'thiosulfate(2-)' = 'tsul[e]', 
  'Methanethiol' = 'ch4s[e]',
  'Hydrogen' = 'h2[e]'
)
metabolite_order <- unname(metabolites_to_plot)

flux_metabolites <- read.csv('../data/2B.csv', check.names = FALSE, row.names = 1) %>%
  mutate(
    metabolite = factor(metabolite, levels = metabolite_order),
    metabolite_name = names(metabolites_to_plot)[match(metabolite, metabolite_order)],
    sample = factor(sample, levels = c("donor", "recipient"))
  )

samples <- unique(flux_metabolites$sample_id)

plot_list <- map2(names(metabolites_to_plot), metabolites_to_plot, function(met_name, met_id) {
  df <- flux_metabolites %>%
    filter(metabolite == met_id)
  
  # compute a y‐position just above the highest point in this facet
  y_max <- max(df$flux, na.rm = TRUE)
  
  ggplot(df, aes(x = sample, y = flux)) +
    geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
    geom_jitter(color = "black", size = 1, width = 0.1, alpha = 0.5) +
    geom_signif(
      comparisons    = list(c("donor", "recipient")),
      test           = "wilcox.test",
      map_signif_level = TRUE,
      textsize       = 3,
      tip_length     = 0.01,
      y_position     = y_max * 1.05
    ) +
    ggtitle(met_name) +
    ylab("flux (mmol/gDW/h)") +
    xlab("") +
    theme_cowplot() +
      coord_cartesian(clip = "off") +
    theme(
      plot.title   = element_text(hjust = 0.5, size = 10),
      panel.grid   = element_blank(),
      strip.background = element_blank(),
          plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt"),
              axis.text = element_text(size = 12),
      axis.title = element_text(size = 14)
    )
})

fig_2c <- ggarrange(
  plotlist = plot_list,
  ncol = 4,
  nrow = 2,
  common.legend = FALSE
)

print(fig_2c)

ggsave(
  "../figures/2c.png",
  fig_2c,
  width = 25,
  height = 12.5,
  dpi = 600,
  bg = "white"
)

```



```{r figure 2D}
scfa_production <- read.csv('../data/2E.csv', check.names = FALSE, row.names = 1) %>%
  mutate(
    clinical_response = factor(clinical_response, 
                               levels = c('responder', 'non-responder', 'donor')),
    time = factor(time, levels = c('T0', 'T6'))
    )

fig_2d <- ggplot(scfa_production, aes(x = time, y = flux)) +
 stat_boxplot(geom ='errorbar', width = 0.2) + 
  geom_boxplot(aes(fill = time), outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.5, height = 0) +
  geom_line(aes(group = base_sample_id),
            colour = "grey70", size = 0.3, alpha = 0.7) +
  # one panel per response group
  facet_wrap(~ clinical_response, scales = 'free_x') +
  scale_fill_brewer(palette = "Set2") +
  scale_colour_brewer(palette = "Set2") +
  labs(x = "Time", y = "flux (mmol/gDW/h)") +
  theme_minimal() +
  theme_cowplot() + 
  theme(
      strip.background = element_blank(),    # remove the strip fill
      strip.text = element_text(size = 14, face = 'bold'),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14)) +
  geom_signif(
    comparisons = list(c("T0","T6")),
    test          = "wilcox.test",
    map_signif_level = TRUE,
    tip_length    = 0.02,
    textsize      = 3.5,
    y_position    = max(scfa_production$flux) * 1.05
  )

print(fig_2d)

ggsave(
  "../figures/2D.png",
  fig_2d,
  width = 15,
  height = 10,
  dpi = 600,
  bg = 'white'
)
```

Assemble figures together
```{r}

fig_2a
fig_2b
fig_2c
fig_2d

top <- plot_grid(fig_2a, fig_2b, rel_widths = c(1.25,1), nrow = 1, labels = c('A', 'B'))
bottom <- plot_grid(fig_2c, fig_2d, rel_widths = c(1.5,1), nrow = 1, labels = c('C', 'D'))

figure2 <- plot_grid(top,bottom, nrow = 2)
figure2
ggsave('../figures/Figure2.pdf', figure2, width = 14, height = 8, dpi = 300, bg = 'white')

```


## Reviewer response: donor vs. recipient community structure (Bray-Curtis PCoA + PERMANOVA)

Addresses the question of whether post-FMT recipients converge on donors in general
or on their own specific donor. Two analyses are run below.
(1) PCoA of Bray-Curtis dissimilarities on relative abundances, with recipients
labeled by a subject number so pre- and post-FMT samples of the same individual can
be traced, and PERMANOVA (adonis2) testing group separation in that space.
(2) Recipient-to-own-donor vs. recipient-to-other-donor Bray-Curtis distances at T0
and T6, which is the comparison that actually distinguishes donor-specific
engraftment from a generic shift toward a "donor-like" community.

```{r pcoa setup}
library(vegan)

raw_pcoa <- read.csv('../data/abundance_raw_subset_filtered.csv', check.names = FALSE) %>%
  column_to_rownames('taxon_id') %>%
  as.matrix()

# relative abundance, samples as rows
rel_pcoa <- t(sweep(raw_pcoa, 2, colSums(raw_pcoa), '/'))

mapping <- read.csv('../../utils/DonorRecipientMapping.csv', check.names = FALSE) %>%
  mutate(
    subject   = str_replace(str_split_fixed(recipient, '_', 2)[,1], '-0$', ''),
    donor_id  = str_split_fixed(donor, '_', 2)[,1]
  ) %>%
  select(subject, donor_id, clinical_response)

# recipient subject numbering used for the point labels (1..n, ordered numerically)
subject_levels <- mapping$subject[order(as.numeric(mapping$subject))]
subject_index  <- setNames(seq_along(subject_levels), subject_levels)

meta_pcoa <- data.frame(sample = rownames(rel_pcoa), stringsAsFactors = FALSE) %>%
  mutate(
    group = case_when(
      str_detect(sample, '^D-')  ~ 'Donor',
      str_detect(sample, '-0$')  ~ 'Recipient pre-FMT (T0)',
      str_detect(sample, '-6$')  ~ 'Recipient post-FMT (T6)'
    ),
    subject   = ifelse(group == 'Donor', NA_character_,
                       str_split_fixed(sample, '-', 2)[,1]),
    rec_label = ifelse(is.na(subject), NA_character_,
                       as.character(subject_index[subject]))
  ) %>%
  left_join(mapping, by = 'subject') %>%
  column_to_rownames('sample')

meta_pcoa <- meta_pcoa[rownames(rel_pcoa), , drop = FALSE]
meta_pcoa$group <- factor(
  meta_pcoa$group,
  levels = c('Donor', 'Recipient pre-FMT (T0)', 'Recipient post-FMT (T6)')
)

bc_dist <- vegdist(rel_pcoa, method = 'bray')
```

```{r PERMANOVA}
set.seed(42)

# global test across the three groups
permanova_global <- adonis2(bc_dist ~ group, data = meta_pcoa,
                            permutations = 9999, by = 'terms')
print(permanova_global)

# dispersion check: PERMANOVA can be driven by unequal within-group spread
bd <- betadisper(bc_dist, meta_pcoa$group)
print(permutest(bd, permutations = 9999))

# pairwise PERMANOVA with BH correction
pairs_to_test <- combn(levels(meta_pcoa$group), 2, simplify = FALSE)

permanova_pairwise <- map_dfr(pairs_to_test, function(pr) {
  keep <- meta_pcoa$group %in% pr
  sub_meta <- droplevels(meta_pcoa[keep, , drop = FALSE])
  sub_dist <- vegdist(rel_pcoa[keep, , drop = FALSE], method = 'bray')
  a <- adonis2(sub_dist ~ group, data = sub_meta,
               permutations = 9999, by = 'terms')
  tibble(
    comparison = paste(pr, collapse = ' vs. '),
    R2 = a$R2[1],
    F_stat = a$F[1],
    p_value = a$`Pr(>F)`[1]
  )
}) %>%
  mutate(p_adj = p.adjust(p_value, method = 'BH'))

print(permanova_pairwise)

write.csv(permanova_pairwise, '../figures/s5_permanova_pairwise.csv', row.names = FALSE)
```

```{r permanova subtitle}
# compact stat strings for the PCoA subtitle
fmt_p <- function(p) ifelse(p < 1e-4, 'p < 1e-4', paste0('p = ', signif(p, 2)))

don_t0 <- permanova_pairwise %>%
  filter(comparison == 'Donor vs. Recipient pre-FMT (T0)')

permanova_subtitle <- paste0(
  'PERMANOVA, donor vs T0: R² = ', sprintf('%.2f', don_t0$R2),
  ', ', fmt_p(don_t0$p_adj)
)

cat(permanova_subtitle, '\n')
```

```{r figure 2E PCoA}
pcoa_res <- cmdscale(bc_dist, k = 4, eig = TRUE)
eig_pos  <- pcoa_res$eig[pcoa_res$eig > 0]
var_expl <- round(100 * pcoa_res$eig[1:2] / sum(eig_pos), 1)

pcoa_df <- as.data.frame(pcoa_res$points[, 1:2])
colnames(pcoa_df) <- c('PCo1', 'PCo2')
pcoa_df <- bind_cols(pcoa_df, meta_pcoa[rownames(pcoa_df), ])

# faint segment linking T0 and T6 of the same recipient
traj_df <- pcoa_df %>%
  filter(!is.na(subject)) %>%
  mutate(time = ifelse(str_detect(group, 'pre'), 'T0', 'T6')) %>%
  select(subject, time, PCo1, PCo2) %>%
  pivot_wider(names_from = time, values_from = c(PCo1, PCo2)) %>%
  filter(!is.na(PCo1_T0), !is.na(PCo1_T6))

group_cols <- c(
  'Donor'                   = '#2C7BB6',
  'Recipient pre-FMT (T0)'  = '#999999',
  'Recipient post-FMT (T6)' = '#D7191C'
)

lab_nudge <- 0.02 * diff(range(pcoa_df$PCo2))

fig_2e <- ggplot(pcoa_df, aes(x = PCo1, y = PCo2)) +
  stat_ellipse(aes(color = group), type = 't', level = 0.8,
               linetype = 'dashed', linewidth = 0.4, show.legend = FALSE) +
  geom_segment(
    data = traj_df,
    aes(x = PCo1_T0, y = PCo2_T0, xend = PCo1_T6, yend = PCo2_T6),
    inherit.aes = FALSE, color = 'grey70', linewidth = 0.25, alpha = 0.7
  ) +
  geom_point(aes(color = group, shape = group), size = 2.2, alpha = 0.9) +
  geom_text(
    aes(label = rec_label, color = group),
    size = 2, vjust = 0, nudge_y = lab_nudge, nudge_x = lab_nudge,
    show.legend = FALSE, na.rm = TRUE
  ) +
  scale_color_manual(values = group_cols, name = '') +
  scale_shape_manual(values = c(17, 16, 16), name = '') +
  labs(
    x = paste0('PCo1 (', var_expl[1], '%)'),
    y = paste0('PCo2 (', var_expl[2], '%)'),
    subtitle = permanova_subtitle
  ) +
  theme_cowplot() +
  theme(
    legend.position = 'bottom',
    legend.text = element_text(size = 9),
    axis.text   = element_text(size = 10),
    axis.title  = element_text(size = 12),
    plot.subtitle = element_text(size = 9, hjust = 0)
  )

print(fig_2e)

ggsave('../figures/s5_pcoa_bray_curtis.pdf', fig_2e,
       width = 6, height = 5.4, dpi = 600, bg = 'white')
ggsave('../figures/s5e_pcoa_bray_curtis.png', fig_2e,
       width = 6, height = 5.4, dpi = 600, bg = 'white')
```

```{r figure 2F own donor vs other donors}
donor_ids <- rownames(rel_pcoa)[str_detect(rownames(rel_pcoa), '^D-')]
bc_mat <- as.matrix(bc_dist)

# Some recipients are assigned to a donor that was not sequenced / not present in
# the abundance table (e.g. D-1), so drop those pairs before indexing bc_mat.
dropped_pairs <- meta_pcoa %>%
  rownames_to_column('sample') %>%
  filter(!is.na(subject), !is.na(donor_id), !(donor_id %in% donor_ids))
if (nrow(dropped_pairs) > 0) {
  message('Dropping ', nrow(dropped_pairs),
          ' recipient sample(s) whose assigned donor is absent from the ',
          'abundance table: ',
          paste(unique(dropped_pairs$donor_id), collapse = ', '))
}

dist_df <- meta_pcoa %>%
  rownames_to_column('sample') %>%
  filter(!is.na(subject), !is.na(donor_id), donor_id %in% donor_ids) %>%
  mutate(time = ifelse(str_detect(group, 'pre'), 'T0', 'T6')) %>%
  rowwise() %>%
  mutate(
    own_donor   = bc_mat[sample, donor_id],
    other_donor = mean(bc_mat[sample, setdiff(donor_ids, donor_id)])
  ) %>%
  ungroup() %>%
  select(sample, subject, time, clinical_response, own_donor, other_donor) %>%
  pivot_longer(c(own_donor, other_donor),
               names_to = 'reference', values_to = 'distance') %>%
  mutate(
    reference = factor(reference,
                       levels = c('own_donor', 'other_donor'),
                       labels = c('Own donor', 'Other donors')),
    time = factor(time, levels = c('T0', 'T6'))
  )

fig_2f <- ggplot(dist_df, aes(x = reference, y = distance)) +
  stat_boxplot(geom = 'errorbar', width = 0.2) +
  geom_boxplot(aes(fill = reference), outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.1, height = 0, size = 1, alpha = 0.5) +
  facet_wrap(~ time) +
  geom_signif(
    comparisons = list(c('Own donor', 'Other donors')),
    test = 'wilcox.test', test.args = list(paired = TRUE),
    map_signif_level = FALSE, textsize = 3, tip_length = 0.01,
    y_position = max(dist_df$distance, na.rm = TRUE) * 1.03
  ) +
  scale_fill_brewer(palette = 'Set2') +
  labs(x = '', y = 'Bray-Curtis distance to donor') +
  theme_cowplot() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = 'bold'),
    axis.text  = element_text(size = 10),
    axis.title = element_text(size = 12)
  )

print(fig_2f)

ggsave('../figures/2f_own_vs_other_donor_distance.pdf', fig_2f,
       width = 5, height = 4, dpi = 600, bg = 'white')

# paired test of whether recipients move closer to their own donor after FMT
own_wide <- dist_df %>%
  filter(reference == 'Own donor') %>%
  select(subject, time, distance) %>%
  pivot_wider(names_from = time, values_from = distance) %>%
  drop_na()

print(wilcox.test(own_wide$T0, own_wide$T6, paired = TRUE))
```
