---
title: "Supplementary Figure 5 - external cohort validation (Li et al., PRJEB46778)"
output: html_document
---

A. Li-native random forest, leave-one-recipient-out, observed vs predicted post-FMT abundance (day 84)
B. Feature importance (impurity, mean +/- SD across LOO folds)
C. Fold-change accuracy, Li-native LOO vs the IBS-trained (Goll) model, pooled across timepoints

Requires `li_et.al_2016_FMT/scripts/01_LiEtAl_dfPrep.py` to have written
`Figure2/data/supp_5/AllMetabolicFeatures.csv`.

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(ggsignif)
library(ranger)
dir.create("../figures", showWarnings = FALSE, recursive = TRUE)

PRIMARY_DAY <- 84   # ~2.8 months, closest Li analogue to the IBS 6-month sample
POST_DAYS   <- c(2, 14, 42, 84)

# fold-change LOO rho achieved by the same model within the IBS cohort
IBS_REFERENCE <- 0.535

rf_features <- c("average_distance", "diversity_ratio",
                 "met_independence", "log10_pre_abund")
rf_formula  <- as.formula(paste("log_fold_change ~", paste(rf_features, collapse = " + ")))

human_palette <- read.csv("../../utils/human_order_palette.csv")
order_colors  <- setNames(human_palette$color, human_palette$order)
named_orders  <- setdiff(names(order_colors), "Other")

tax <- read.csv("../../utils/full_taxonomy.csv") %>%
  select(taxon_id, order) %>%
  distinct()

clean_df <- function(d) {
  d %>%
    filter(if_all(all_of(c(rf_features, "log_fold_change",
                           "pre_abundance", "post_abundance")), ~ !is.na(.x))) %>%
    filter(is.finite(log_fold_change), is.finite(log10_pre_abund))
}

li_all <- read.csv("../data/supp_5/AllMetabolicFeatures.csv") %>%
  mutate(comparison = paste(donor_subject, "vs", recipient))

AllMetabolicFeatures <- clean_df(li_all %>% filter(timepoint_day == PRIMARY_DAY))

rf_mod <- readRDS("../../Figure3/processed_data/rf_model_IBS.rds")
```

Panel A. Leave-one-recipient-out CV within the Li cohort, day 84.

```{r}
set.seed(123)
recipients <- unique(AllMetabolicFeatures$recipient)

df_rf <- AllMetabolicFeatures
df_rf$loo_pred_fc <- NA_real_
imp_matrix <- matrix(NA, nrow = length(recipients), ncol = length(rf_features),
                     dimnames = list(recipients, rf_features))

for (i in seq_along(recipients)) {
  rec <- recipients[i]
  mod <- ranger(rf_formula,
                data          = filter(df_rf, recipient != rec),
                num.trees     = 100,
                mtry          = 3,
                min.node.size = 2,
                importance    = "impurity")
  df_rf$loo_pred_fc[df_rf$recipient == rec] <-
    predict(mod, data = filter(df_rf, recipient == rec))$predictions
  imp_matrix[i, ] <- mod$variable.importance
}

df_rf <- df_rf %>% mutate(loo_pred_abundance = pre_abundance * 2 ^ loo_pred_fc)

loo_rho <- df_rf %>%
  group_by(recipient) %>%
  summarise(rho = cor(post_abundance, loo_pred_abundance,
                      method = "spearman", use = "complete.obs"), .groups = "drop")

cat("Li-native LOO rho (abundance):", round(median(loo_rho$rho), 2),
    "+-", round(sd(loo_rho$rho), 2), "\n")

df_plot <- df_rf %>%
  left_join(tax, by = c("taxon" = "taxon_id")) %>%
  mutate(order_plot = if_else(order %in% named_orders, order, "Other"),
         order_plot = factor(order_plot, levels = c(named_orders, "Other")))

panelA <- ggplot(df_plot, aes(x = log10(post_abundance + 1e-6),
                              y = log10(loo_pred_abundance + 1e-6))) +
  geom_point(aes(color = order_plot), alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  scale_colour_manual(values = order_colors, name = "Order") +
  annotate("text",
           x = min(log10(df_rf$post_abundance + 1e-6), na.rm = TRUE) + 0.5,
           y = max(log10(df_rf$loo_pred_abundance + 1e-6), na.rm = TRUE) - 0.3,
           label = paste0("rho == ", round(median(loo_rho$rho), 2)),
           parse = TRUE, hjust = 0, size = 4) +
  labs(x = expression(Observed~log[10](Post-FMT~Abundance)),
       y = expression(Predicted~log[10](Post-FMT~Abundance))) +
  theme_cowplot()

panelA
```

Panel B. Impurity importance across LOO folds.

```{r}
imp_df <- data.frame(Feature    = colnames(imp_matrix),
                     Importance = colMeans(imp_matrix),
                     SD         = apply(imp_matrix, 2, sd)) %>%
  mutate(Feature = recode(Feature,
                          average_distance = "Metabolic\nDistance",
                          met_independence = "Metabolic\nIndependence",
                          log10_pre_abund  = "Pre-FMT\nAbundance",
                          diversity_ratio  = "Diversity\nRatio")) %>%
  arrange(desc(Importance)) %>%
  mutate(Feature = factor(Feature, levels = Feature))

panelB <- ggplot(imp_df, aes(x = Feature, y = Importance)) +
  geom_col(fill = "steelblue", color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = pmax(Importance - SD, 0), ymax = Importance + SD),
                width = 0.2) +
  labs(x = NULL, y = "Impurity Importance\n(mean +/- SD across LOO folds)") +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

panelB
print(imp_df)
```

Panel C. Performance of the IBS-trained (Goll) model applied to this cohort
without retraining, across all four post-FMT timepoints. Scored on log2
fold-change rather than predicted abundance, because predicted abundance is the
product of pre-FMT abundance and predicted fold-change and is therefore
dominated by the pre-FMT term. Dashed line is the fold-change accuracy the same
model achieves within the IBS cohort by leave-one-recipient-out CV.

```{r}
set.seed(123)

goll_rho <- map_dfr(POST_DAYS, function(day) {
  d <- clean_df(li_all %>% filter(timepoint_day == day))
  d$goll_pred_fc <- predict(rf_mod, d)$predictions

  d %>%
    group_by(recipient, fmt_type) %>%
    summarise(rho = cor(log_fold_change, goll_pred_fc, method = "spearman"),
              .groups = "drop") %>%
    mutate(day = day)
}) %>%
  mutate(day = factor(day, levels = POST_DAYS))

goll_rho %>%
  group_by(day) %>%
  summarise(median_rho = round(median(rho), 3), sd = round(sd(rho), 3),
            .groups = "drop") %>%
  print()

cat("pooled across timepoints: median rho =",
    round(median(goll_rho$rho), 3), "+-", round(sd(goll_rho$rho), 3), "\n")

panelC <- ggplot(goll_rho, aes(x = day, y = rho)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_hline(yintercept = IBS_REFERENCE, linetype = "dashed",
             color = "firebrick", linewidth = 0.5) +
  stat_boxplot(geom = "errorbar", width = 0.2) +
  geom_boxplot(width = 0.5, fill = "#F1E201FF", outlier.shape = NA) +
  geom_jitter(aes(shape = fmt_type), height = 0, width = 0.12,
              alpha = 0.8, size = 1.8) +
  annotate("text", x = 0.55, y = IBS_REFERENCE + 0.05,
           label = "IBS cohort", color = "firebrick", size = 3, hjust = 0) +
  scale_shape_manual(values = c(Allogenic = 16, Autologous = 1), name = NULL) +
  labs(x = "Days post-FMT", y = expression(Fold-change~Spearman~rho)) +
  theme_cowplot() +
  theme(legend.position = "bottom")

panelC
```

Alternative Panel C, if you would rather mirror the layout of panel A. This is
the Goll model scored on the same axes as A, observed versus predicted post-FMT
abundance at day 84. Note that it lands at rho = 0.76 against panel A's 0.78,
so a reader comparing the two panels would conclude the transfer worked, which
is the opposite of the point the text makes.

```{r, eval=FALSE}
goll_day84 <- AllMetabolicFeatures %>%
  mutate(goll_pred_fc        = predict(rf_mod, .)$predictions,
         goll_pred_abundance = pre_abundance * 2 ^ goll_pred_fc) %>%
  left_join(tax, by = c("taxon" = "taxon_id")) %>%
  mutate(order_plot = if_else(order %in% named_orders, order, "Other"),
         order_plot = factor(order_plot, levels = c(named_orders, "Other")))

goll_abd_rho <- goll_day84 %>%
  group_by(recipient) %>%
  summarise(rho = cor(post_abundance, goll_pred_abundance,
                      method = "spearman", use = "complete.obs"), .groups = "drop")

panelC_alt <- ggplot(goll_day84, aes(x = log10(post_abundance + 1e-6),
                                     y = log10(goll_pred_abundance + 1e-6))) +
  geom_point(aes(color = order_plot), alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  scale_colour_manual(values = order_colors, name = "Order") +
  annotate("text",
           x = min(log10(goll_day84$post_abundance + 1e-6), na.rm = TRUE) + 0.5,
           y = max(log10(goll_day84$goll_pred_abundance + 1e-6), na.rm = TRUE) - 0.3,
           label = paste0("rho == ", round(median(goll_abd_rho$rho), 2)),
           parse = TRUE, hjust = 0, size = 4) +
  labs(x = expression(Observed~log[10](Post-FMT~Abundance)),
       y = expression(IBS-trained~model~log[10](Predicted))) +
  theme_cowplot()

panelC_alt
```

Assemble.

```{r}
shared_legend <- get_legend(panelA + theme(legend.position = "right"))

fig <- plot_grid(
  panelA + theme(legend.position = "none"),
  panelB,
  panelC,
  labels = "AUTO", nrow = 1, rel_widths = c(1, 0.8, 0.75)
)

fig_final <- plot_grid(fig, shared_legend, nrow = 1, rel_widths = c(1, 0.2))
fig_final

ggsave("../figures/supplementary_figure_5.pdf", fig_final, width = 12, height = 3.5)
ggsave("../figures/supplementary_figure_5.png", fig_final, width = 12, height = 3.5, bg = "white")
```
