---
title: "Metabolic Niche Analysis"
subtitle: "PCoA + Hypervolume of MICOM exchange fluxes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}

setwd("/Users/guillaumeurtecho/Dropbox/UrtechoLab/Projects/2024/FMT_Metabolic_models_ZZ/20260312_NicheRangeAnalysis_JCrTE/scripts")

library(tidyverse)
library(proxy)
library(ape)
library(ggplot2)
library(ggrepel)
library(hypervolume)
library(viridis)
library(patchwork)
library(cluster)     # pam(), silhouette()
library(factoextra)  # fviz_silhouette()
library(cowplot)

flux <- read.csv("../input/exchanges_d0_high_fiber.csv") %>%
  filter(!grepl("medium", taxon, ignore.case = TRUE)) #remove medium

#scale by abundance?
flux <- flux %>%
  mutate(flux = (flux / abundance) * median(abundance))

tax <- read.csv("../ref/assembly.taxonomy_parsed_filled.csv")


# ── Load shared palette ───────────────────────────────────────────────────
palette_df <- read.csv("../../FMT/utils/colorPalette.csv", row.names = 1)
order_colors <- setNames(palette_df$x, rownames(palette_df))
named_orders <- setdiff(names(order_colors), "Other")

#color palette
make_top_order_palette <- function(data, tax_col = "order", n_top = 12,
                                   other_label = "Other", other_colour = "grey75") {
  top_orders <- data %>%
    count(.data[[tax_col]], sort = TRUE) %>%
    slice_head(n = n_top) %>%
    pull(.data[[tax_col]])

  colours <- setNames(
    c(RColorBrewer::brewer.pal(n_top, "Set3"), other_colour),
    c(top_orders, other_label)
  )
  list(top_orders = top_orders, colours = colours, other_label = other_label)
}


```

## 1 · Build & Preprocess the Flux Matrix

Rows are `taxon__sample_id` combinations; columns are `metabolite_direction`
(e.g. `cit_e_import`, `cit_e_export`) with absolute flux values. Zeros receive
a data-adaptive pseudocount (1% of the minimum observed non-zero flux) before
log transformation. 

NO MORE ###No abundance weighting is applied — each taxon is treated
##as an equal member of its community.

```{r build-matrix}
# ── Wide matrix ───────────────────────────────────────────────────────────────
flux_wide <- flux %>%
  mutate(
    taxon_sample = paste(taxon, sample_id, sep = "__"),
    feature      = paste(metabolite, direction, sep = "_"),
    flux_abs     = abs(flux)
  ) %>%
  pivot_wider(
    id_cols     = c(taxon_sample, sample_id),
    names_from  = feature,
    values_from = flux_abs,
    values_fill = 0,
    values_fn   = mean   # average if duplicate reaction rows exist
  )

meta <- flux_wide %>% select(taxon_sample, sample_id)
mat  <- flux_wide %>% select(-taxon_sample, -sample_id) %>% as.matrix()
rownames(mat) <- flux_wide$taxon_sample

message(sprintf("Matrix: %d taxon×sample rows  ×  %d features", nrow(mat), ncol(mat)))
message(sprintf("Sparsity: %.1f%% zeros", mean(mat == 0) * 100))

# ── Pseudocount + log transform ───────────────────────────────────────────────
pseudocount <- min(mat[mat > 0], na.rm = TRUE) * 0.01
message(sprintf("Pseudocount: %.2e", pseudocount))

mat_log <- log(mat + pseudocount)
```

## 2 · Prevalence Filtering

Keep only metabolite-direction features exchanged by ≥ `prev_threshold` of
all taxa×sample rows globally. The diagnostic plot shows the full prevalence
distribution — adjust the threshold based on where the distribution drops off.

```{r filter-features, fig.height=4}
# ── Set threshold ─────────────────────────────────────────────────────────────
prev_threshold <- 0.01   # <-- adjust: proportion of taxa that must exchange
                          #     a metabolite for it to be retained

# ── Prevalence per feature ────────────────────────────────────────────────────
prevalence <- colMeans(mat > 0)

prev_df <- tibble(feature = names(prevalence), prevalence = prevalence)
n_pass  <- sum(prevalence >= prev_threshold)

# ── Diagnostic plot ───────────────────────────────────────────────────────────
p_prev <- ggplot(prev_df, aes(x = prevalence)) +
  geom_histogram(bins = 60, fill = "steelblue", alpha = 0.8, colour = "white") +
  geom_vline(xintercept = prev_threshold, colour = "firebrick",
             linetype = "dashed", linewidth = 0.9) +
  annotate("text",
           x = prev_threshold + 0.01, y = Inf, vjust = 1.5, hjust = 0,
           label = sprintf("threshold = %.0f%%  |  %d / %d features retained",
                           prev_threshold * 100, n_pass, ncol(mat)),
           colour = "firebrick", size = 3.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title    = "Feature prevalence distribution (pre-filter)",
    subtitle = "Each feature = one metabolite × direction pair",
    x        = "Prevalence (proportion of taxa with non-zero flux)",
    y        = "Number of features"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p_prev)
message(sprintf("Retaining %d / %d features (>= %.0f%% prevalence)",
                n_pass, ncol(mat), prev_threshold * 100))

# ── Apply filter & overwrite mat_log for all downstream steps ─────────────────
keep_features <- names(prevalence[prevalence >= prev_threshold])
mat_filt      <- mat[, keep_features]
mat_log       <- log(mat_filt + pseudocount)

message(sprintf("Matrix after filtering: %d rows x %d features  |  sparsity %.1f%%",
                nrow(mat_log), ncol(mat_log), mean(mat_filt == 0) * 100))
```

## 3 · Cosine Distance & PCoA

Cosine distance (`1 − cosine similarity`) captures differences in niche
*shape* independently of overall flux magnitude. The Cailliez correction
handles any negative eigenvalues that cosine distances can occasionally
produce.

```{r pcoa, fig.height=9}
# ── Cosine distance ───────────────────────────────────────────────────────────
cos_dist <- as.dist(1 - as.matrix(proxy::simil(mat_log, method = "cosine")))

# ── PCoA ──────────────────────────────────────────────────────────────────────
pcoa_res <- ape::pcoa(cos_dist, correction = "cailliez")
var_exp  <- round(pcoa_res$values$Rel_corr_eig * 100, 2)

message(sprintf("PCo1: %.1f%%  PCo2: %.1f%%  PCo3: %.1f%%",
                var_exp[1], var_exp[2], var_exp[3]))

# ── Tidy coordinates + centroids ──────────────────────────────────────────────
pcoa_coords <- as.data.frame(pcoa_res$vectors.cor) %>%
  rownames_to_column("taxon_sample") %>%
  left_join(meta, by = "taxon_sample")

centroids <- pcoa_coords %>%
  group_by(sample_id) %>%
  summarise(across(starts_with("Axis."), mean), .groups = "drop")

sample_colours <- setNames(
  viridis::turbo(n_distinct(meta$sample_id)),
  sort(unique(meta$sample_id))
)




# ── PCoA plot ─────────────────────────────────────────────────────────────────
p_pcoa <- ggplot(pcoa_coords, aes(Axis.1, Axis.2, colour = sample_id)) +
  stat_ellipse(aes(group = sample_id), type = "norm",
               linetype = "dashed", alpha = 0.5, linewidth = 0.7) +
  geom_point(alpha = 0.55, size = 1.8) +
  geom_point(data = centroids, shape = 3, size = 6, stroke = 2.5) +
  ggrepel::geom_label_repel(
    data = centroids, aes(label = sample_id),
    size = 3.8, fontface = "bold", show.legend = FALSE,
    box.padding = 0.4, seed = 42
  ) +
  scale_colour_manual(values = sample_colours, name = "Sample group") +
  labs(
    title    = "Metabolic Niche Space — PCoA (Cosine Distance)",
    subtitle = "Each point = one taxon in one sample  |  ✚ = community centroid  |  Dashed = 95% ellipse",
    x = paste0("PCo1 (", var_exp[1], "%)"),
    y = paste0("PCo2 (", var_exp[2], "%)")
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey40"))

# ── Scree plot ────────────────────────────────────────────────────────────────
scree_df <- tibble(
  axis   = seq_len(min(20, length(var_exp))),
  var    = var_exp[seq_len(min(20, length(var_exp)))],
  cumvar = cumsum(var_exp)[seq_len(min(20, length(var_exp)))]
)

p_scree <- ggplot(scree_df, aes(x = axis)) +
  geom_col(aes(y = var), fill = "steelblue", alpha = 0.8) +
  geom_line(aes(y = cumvar), colour = "firebrick", linewidth = 1) +
  geom_point(aes(y = cumvar), colour = "firebrick", size = 2) +
  geom_hline(yintercept = 80, linetype = "dashed", colour = "grey40") +
  annotate("text", x = max(scree_df$axis) * 0.75, y = 82,
           label = "80% threshold", colour = "grey40", size = 3.5) +
  scale_x_continuous(breaks = scree_df$axis) +
  labs(title = "PCoA Scree Plot", x = "PCoA Axis", y = "Variance explained (%)") +
  theme_bw(base_size = 12)

(p_pcoa) / (p_scree)

#can also get boxplots, but these aren't significant!



```

## 4 · Niche Count (PAM Clustering)

For each sample group, PAM clustering is run on the subset of the cosine
distance matrix at k = 2 through `k_max`. The optimal k is chosen by the
highest mean silhouette width — a score near 1 means taxa are tightly assigned
to their niche, near 0 means the boundary is ambiguous. Cluster assignments
are then projected back onto the PCoA for visual inspection.

```{r niche-count, fig.height=6}
# ── Parameters ────────────────────────────────────────────────────────────────
k_max <- 50   # <-- maximum number of niches to test per sample

# ── PAM + silhouette per sample group ─────────────────────────────────────────
cos_dist_mat        <- as.matrix(cos_dist)
niche_count_results <- list()
cluster_assignments <- list()

for (s in sort(unique(meta$sample_id))) {
  idx    <- meta$taxon_sample[meta$sample_id == s]
  n_taxa <- length(idx)
  submat <- cos_dist_mat[idx, idx]

  k_range <- 2:min(k_max, n_taxa - 1)
  if (length(k_range) < 1) {
    warning(sprintf("Sample '%s': too few taxa to cluster (%d)", s, n_taxa))
    next
  }

  # Run PAM at each k, record mean silhouette width
  sil_scores <- sapply(k_range, function(k) {
    pam_fit <- pam(as.dist(submat), k = k, diss = TRUE)
    mean(silhouette(pam_fit$clustering, as.dist(submat))[, "sil_width"])
  })

  best_k   <- k_range[which.max(sil_scores)]
  best_sil <- max(sil_scores)
  message(sprintf("Sample %s: optimal k = %d  (mean silhouette = %.3f)",
                  s, best_k, best_sil))

  pam_best <- pam(as.dist(submat), k = best_k, diss = TRUE)

  niche_count_results[[s]] <- tibble(
    sample_id       = s,
    n_taxa          = n_taxa,
    n_niches        = best_k,
    mean_silhouette = best_sil
  )

  cluster_assignments[[s]] <- tibble(
    taxon_sample = idx,
    sample_id    = s,
    niche        = factor(pam_best$clustering)
  )

  # Per-sample silhouette bar plot
  sil_df <- as.data.frame(silhouette(pam_best$clustering, as.dist(submat))) %>%
    arrange(cluster, sil_width) %>%
    mutate(rank = row_number(), cluster = factor(cluster))

  p_sil <- ggplot(sil_df, aes(x = rank, y = sil_width, fill = cluster)) +
    geom_col(width = 1, colour = NA) +
    geom_hline(yintercept = best_sil, linetype = "dashed",
               colour = "grey30", linewidth = 0.7) +
    scale_fill_viridis_d(option = "turbo", name = "Niche") +
    scale_y_continuous(limits = c(NA, 1)) +
    labs(
      title    = sprintf("Sample %s — silhouette (k = %d)", s, best_k),
      subtitle = sprintf("Mean silhouette width = %.3f", best_sil),
      x = "Taxa (sorted within cluster)", y = "Silhouette width"
    ) +
    theme_bw(base_size = 11) +
    theme(plot.title    = element_text(face = "bold"),
          plot.subtitle = element_text(size = 9, colour = "grey40"))

  print(p_sil)
}

# ── Summary table ─────────────────────────────────────────────────────────────
niche_summary <- bind_rows(niche_count_results)
print(niche_summary)

# ── Project cluster assignments onto PCoA ─────────────────────────────────────
pcoa_clust <- pcoa_coords %>%
  left_join(bind_rows(cluster_assignments), by = c("taxon_sample", "sample_id"))

pcoa_clust <- pcoa_clust %>%
  mutate(order_plot = if_else(order %in% named_orders, order, "Other"),
         order_plot = factor(order_plot, levels = c(named_orders, "Other")))

# ── PCoA faceted by sample, coloured by niche cluster ─────────────────────────
p_pcoa_clust <- ggplot(pcoa_clust, aes(Axis.1, Axis.2, colour = niche)) +
  geom_point(alpha = 0.7, size = 2) +
  facet_wrap(~ sample_id, nrow = 2) +
  scale_colour_manual(values = order_colors, name = "Order") +
  scale_fill_manual(values = order_colors, name = "Order") +  labs(
    title    = "Metabolic Niches in PCoA Space",
    subtitle = "Each panel = one sample  |  colour = PAM niche cluster (optimal k)",
    x = paste0("PCo1 (", var_exp[1], "%)"),
    y = paste0("PCo2 (", var_exp[2], "%)")
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey40"),
        strip.text    = element_text(face = "bold"))

print(p_pcoa_clust)


```

#orr, plot by taxonomy!
```{r}

# make_top_order_palette <- function(data, tax_col = "order", n_top = 12,
#                                    other_label = "Other", other_colour = "grey75") {
#   top_orders <- data %>%
#     count(.data[[tax_col]], sort = TRUE) %>%
#     slice_head(n = n_top) %>%
#     pull(.data[[tax_col]])
# 
#   colours <- setNames(
#     c(RColorBrewer::brewer.pal(n_top, "Set3"), other_colour),
#     c(top_orders, other_label)
#   )
#   list(top_orders = top_orders, colours = colours, other_label = other_label)
# }

# 1. Build pcoa_clust with taxonomy first
pcoa_clust <- pcoa_coords %>%
  left_join(bind_rows(cluster_assignments), by = c("taxon_sample", "sample_id")) %>%
  mutate(assembly = sub("__.*$", "", taxon_sample)) %>%
  left_join(tax, by = "assembly")

# 2. Then build palette from it
order_pal <- make_top_order_palette(pcoa_clust)

# 3. Then add order_plot
pcoa_clust <- pcoa_clust %>%
  mutate(order_plot = if_else(order %in% order_pal$top_orders, order, order_pal$other_label),
         order_plot = factor(order_plot, levels = c(order_pal$top_orders, order_pal$other_label)))

# 4. Plot
p_pcoa_clust <- ggplot(pcoa_clust, aes(Axis.1, Axis.2, colour = order_plot, fill = order_plot)) +
  geom_point(shape = 21, size = 2.5, alpha = 0.85, stroke = 0.6) +
  facet_wrap(~ sample_id, nrow = 2) +
  scale_colour_manual(values = order_pal$colours, name = "Order") +
  scale_fill_manual(values = order_pal$colours, name = "Order") +
  labs(
   # title    = "Metabolic Niches in PCoA Space",
  #  subtitle = "Each panel = one sample  |  colour = taxonomic order",
    x = paste0("PCo1 (", var_exp[1], "%)"),
    y = paste0("PCo2 (", var_exp[2], "%)")
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(size = 9, colour = "grey40"),
        strip.text      = element_text(face = "bold"),
        strip.background = element_blank(),
        legend.text     = element_text(size = 10, face = 'italic'),
        legend.key.size = unit(0.4, "cm"),
        axis.text = element_blank())

print(p_pcoa_clust)



ggsave("../figs/niche_tax_clust.pdf", width = 5, height = 3.5)

```

## 5 · Grid Occupancy & Density

A shared global grid is overlaid on PCo1 × PCo2 space (covering all samples
together) at two resolutions: 10×10 and 20×20. For each sample and each
resolution, two metrics are computed: **occupancy** — the number of cells
containing at least one taxon (niche diversity), and **mean density within
occupied cells** — the average taxa count per non-empty cell (niche packing).
The heatmaps use a shared colour scale across samples so cross-group comparisons
are direct.

```{r grid-occupancy, fig.height=10}
#grid_sizes <- c(10, 20)
grid_sizes <- c(20)

ax1_range <- range(pcoa_coords$Axis.1)
ax2_range <- range(pcoa_coords$Axis.2)

pad <- 0.01
ax1_range <- ax1_range + c(-pad, pad) * diff(ax1_range)
ax2_range <- ax2_range + c(-pad, pad) * diff(ax2_range)

grid_results <- list()
heat_plots   <- list()

for (g in grid_sizes) {
  breaks1 <- seq(ax1_range[1], ax1_range[2], length.out = g + 1)
  breaks2 <- seq(ax2_range[1], ax2_range[2], length.out = g + 1)

  pcoa_binned <- pcoa_coords %>%
    mutate(
      bin1 = cut(Axis.1, breaks = breaks1, include.lowest = TRUE, labels = FALSE),
      bin2 = cut(Axis.2, breaks = breaks2, include.lowest = TRUE, labels = FALSE)
    )

  cell_counts <- pcoa_binned %>%
    group_by(sample_id, bin1, bin2) %>%
    summarise(n = n(), .groups = "drop")

  grid_summary <- cell_counts %>%
    group_by(sample_id) %>%
    summarise(
      occupancy    = n(),
      total_cells  = g * g,
      pct_occupied = occupancy / total_cells * 100,
      mean_density = mean(n),
      max_density  = max(n),
      .groups = "drop"
    ) %>%
    mutate(grid = paste0(g, "×", g))

  full_grid <- expand.grid(
    sample_id = unique(pcoa_coords$sample_id),
    bin1      = seq_len(g),
    bin2      = seq_len(g)
  ) %>%
    left_join(cell_counts, by = c("sample_id", "bin1", "bin2")) %>%
    replace_na(list(n = 0)) %>%
    mutate(
      x_mid = breaks1[bin1] + diff(breaks1[1:2]) / 2,
      y_mid = breaks2[bin2] + diff(breaks2[1:2]) / 2
    )

  grid_results[[as.character(g)]] <- list(
    summary     = grid_summary,
    cell_counts = cell_counts %>% mutate(grid = paste0(g, "×", g)),
    full_grid   = full_grid %>% mutate(grid = paste0(g, "×", g)),
    breaks1     = breaks1,
    breaks2     = breaks2
  )

  cell_w <- diff(breaks1[1:2])
  cell_h <- diff(breaks2[1:2])

p_heat <- ggplot(full_grid,
                   aes(x = x_mid, y = y_mid, fill = n)) +
    geom_tile(aes(width = cell_w, height = cell_h), colour = "white", linewidth = 0.2) +
    facet_wrap(~ sample_id, nrow = 2) +
    scale_fill_viridis_c(
      option    = "magma",
      direction = -1,
      name      = "Taxa\nper cell",
      limits    = c(0, max(full_grid$n))
    ) +
 #   geom_text(
#      data = grid_summary,
#      aes(x = mean(ax1_range), y = ax2_range[2] * 0.92,
#          label = sprintf("occ: %d/%d  |  density: %.1f",
#                          occupancy, total_cells, mean_density)),
#      inherit.aes = FALSE,
#      size = 3, colour = "grey20", fontface = "bold"
#    ) +
    coord_fixed() +
    labs(
    #  title    = sprintf("Grid occupancy — %d×%d (shared global grid)", g, g),
   #   subtitle = "occ = occupied cells  |  density = mean taxa per occupied cell",
      x = paste0("PCo1 (", var_exp[1], "%)"),
      y = paste0("PCo2 (", var_exp[2], "%)")
    ) +
    theme_bw(base_size = 12) +
    theme(plot.title    = element_text(face = "bold"),
          plot.subtitle = element_text(size = 9, colour = "grey40"),
          strip.text    = element_text(face = "bold"),
          strip.background = element_blank(),
          panel.grid    = element_blank(),
          axis.text = element_blank())

  heat_plots[[paste0("p_heat_", g)]] <- p_heat
  print(p_heat)
}

ggsave("../figs/niche_cell_clust.pdf", width = 4.25, height = 3.5)


# ── Combined summary ──────────────────────────────────────────────────────────
grid_summary_all <- bind_rows(lapply(grid_results, `[[`, "summary")) %>%
  select(grid, sample_id, occupancy, total_cells, pct_occupied, mean_density, max_density) %>%
  arrange(grid, sample_id)

print(grid_summary_all)

# ── Occupancy bar chart (raw cell count) ──────────────────────────────────────
p_occ <- ggplot(grid_summary_all,
                aes(x = sample_id, y = occupancy, fill = sample_id)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%d", occupancy)),
            vjust = -0.4, size = 3.2) +
  facet_wrap(~ grid) +
  scale_fill_manual(values = sample_colours) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Niche space occupancy by community",
    subtitle = "Number of global grid cells containing ≥1 taxon",
    x = NULL, y = "Occupied cells (count)"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey40"),
        strip.text    = element_text(face = "bold"),
        strip.background = element_blank())

p_dens <- ggplot(grid_summary_all,
                 aes(x = sample_id, y = mean_density, fill = sample_id)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f", mean_density)),
            vjust = -0.4, size = 3.2) +
  facet_wrap(~ grid) +
  scale_fill_manual(values = sample_colours) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Niche density by community",
    subtitle = "Mean taxa per occupied grid cell",
    x = NULL, y = "Mean taxa per\noccupied cell"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey40"),
        strip.text    = element_text(face = "bold"),
        strip.background = element_blank())

print(p_occ / p_dens)



#alternative boxplot:

# ── Per-cell density data (one row per occupied cell per sample per grid) ──────
cell_density_all <- bind_rows(lapply(grid_results, `[[`, "full_grid")) %>%
  select(sample_id, n, grid, bin1, bin2) %>%
  group_by(grid, bin1, bin2) %>%
  filter(any(n > 0)) %>%          # keep cell only if occupied in at least one sample
  ungroup() %>%
  select(sample_id, n, grid)

# All pairwise comparisons for geom_signif
sample_levels  <- sort(unique(meta$sample_id))
signif_pairs   <- combn(sample_levels, 2, simplify = FALSE)

p_dens_boxplot <- ggplot(cell_density_all,
                 aes(x = sample_id, y = n, fill = sample_id)) +
  stat_boxplot(geom = "errorbar", width = 0.25, colour = "grey30") +
  geom_boxplot(
    width = 0.45,
    outlier.shape = NA,   # outliers shown via jitter instead
    show.legend = FALSE
  ) +
  geom_jitter(
    width  = 0.15,
    height = 0,
    size   = 1.5,
    alpha  = 0.5,
    show.legend = FALSE
  ) +
  ggsignif::geom_signif(
    comparisons  = signif_pairs,
    test         = "wilcox.test",
    step_increase = 0.1,
    tip_length   = 0.01,
    textsize     = 3,
    colour       = "grey30",
    map_signif_level = T
  ) +
  facet_wrap(~ grid, scales = 'free') +
  scale_fill_manual(values = sample_colours) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
  labs(
    title    = "Niche density by community",
    subtitle = "Taxa per occupied grid cell  |  error bars = mean ± SE  |  Wilcoxon p-values",
    x = NULL, y = "Taxa per occupied cell"
  ) +
  theme_cowplot() +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey40"),
        strip.text    = element_text(face = "bold"),
        strip.background = element_blank())

p_dens_boxplot
ggsave("../figs/OccupiedCells.pdf", width = 1.5, height = 3.5)

```

Let's marginal bar plot?
```{r}
p_occ_side <- ggplot(grid_summary,
                     aes(y = occupancy, x = sample_id, fill = sample_id)) +
  geom_col(width = 0.3, show.legend = FALSE) +
  geom_text(aes(label = occupancy), vjust = -1, size = 3.5) +
  scale_fill_manual(values = sample_colours) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.4))) +
  facet_wrap(~sample_id, scales = 'free_x') +
  labs(y = "Occupied\ncells", x = NULL) +
  theme_cowplot() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.background = element_blank())

p_occ_side

```




## 7 · Hypervolume & Niche Breadth

The number of PCoA axes needed to explain ≥80% of variance is determined
automatically. If that exceeds 20 axes, the hypervolume becomes computationally
unreliable in high dimensions, so the analysis falls back to **mean pairwise
cosine distance** within each sample group — a direct, interpretable measure of
how spread out each community's taxa are in niche space.

```{r hypervolume, fig.height=6}
# ── Determine axes & choose method ───────────────────────────────────────────
cumvar       <- cumsum(var_exp)
n_axes_80    <- which(cumvar >= 80)[1]
n_hv_axes    <- min(n_axes_80, 50)          # hard cap: hypervolume unreliable >~20D
use_fallback <- is.na(n_axes_80) | n_hv_axes > 20

message(sprintf("Axes for ≥80%% variance: %s  |  Method: %s",
                ifelse(is.na(n_axes_80), ">all", n_axes_80),
                ifelse(use_fallback, "mean pairwise cosine distance", 
                       paste0("hypervolume (", n_hv_axes, "D)"))))

# ── Niche breadth calculation ─────────────────────────────────────────────────
if (use_fallback) {
  # Mean pairwise cosine distance within each sample group.
  # Higher = taxa occupy more spread-out, distinct metabolic niches.
  niche_breadth <- meta %>%
    group_by(sample_id) %>%
    group_map(function(d, g) {
      idx  <- d$taxon_sample
      subm <- as.matrix(cos_dist)[idx, idx]
      tibble(
        sample_id     = g$sample_id,
        niche_breadth = mean(subm[upper.tri(subm)]),
        n_taxa        = nrow(d),
        metric        = "mean pairwise cosine distance"
      )
    }) %>%
    bind_rows()

} else {
  # Gaussian hypervolume on top n_hv_axes PCoA axes
  hv_axis_cols <- paste0("Axis.", seq_len(n_hv_axes))

  hv_list <- list()
  for (s in sort(unique(meta$sample_id))) {
    coords_s <- pcoa_coords %>%
      filter(sample_id == s) %>%
      select(all_of(hv_axis_cols)) %>%
      as.matrix()

    if (nrow(coords_s) < 5) {
      warning(sprintf("Sample '%s': only %d taxa, skipping (min=5)", s, nrow(coords_s)))
      next
    }
    message(sprintf("Fitting hypervolume — %s (%d taxa)", s, nrow(coords_s)))
    hv_list[[s]] <- hypervolume_gaussian(coords_s, name = s, verbose = FALSE)
  }

  # Pairwise Sørensen overlaps
  n_s         <- length(hv_list)
  overlap_mat <- matrix(NA, n_s, n_s, dimnames = list(names(hv_list), names(hv_list)))
  diag(overlap_mat) <- 1
  if (n_s >= 2) {
    for (i in seq_len(n_s - 1)) {
      for (j in (i + 1):n_s) {
        pair     <- hypervolume_set(hv_list[[i]], hv_list[[j]], check.memory = FALSE)
        sorensen <- pair@HVList$Intersection@Volume * 2 /
                    (hv_list[[i]]@Volume + hv_list[[j]]@Volume)
        overlap_mat[i, j] <- overlap_mat[j, i] <- sorensen
      }
    }
    message("Pairwise Sørensen overlaps:")
    print(round(overlap_mat, 3))
  }

  niche_breadth <- tibble(
    sample_id     = names(hv_list),
    niche_breadth = sapply(hv_list, \(h) h@Volume),
    n_taxa        = sapply(names(hv_list), \(s) sum(meta$sample_id == s)),
    metric        = paste0("hypervolume (", n_hv_axes, "D)")
  )
}

print(niche_breadth)

# ── Niche breadth bar plot ────────────────────────────────────────────────────
metric_label <- unique(niche_breadth$metric)

p_hv <- ggplot(niche_breadth,
               aes(reorder(sample_id, -niche_breadth), niche_breadth, fill = sample_id)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0("n=", n_taxa)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = sample_colours) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Metabolic distance within by Community",
    subtitle = paste0("Metric: ", metric_label),
    x = NULL, y = metric_label
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_hv)
```

## 8 · Export Results

```{r export}
ggsave("../figs/niche_pcoa.pdf",         plot = p_pcoa,                    width = 5, height = 3.5)
ggsave("../figs/niche_clusters.pdf",     plot = p_pcoa_clust,              width = 6, height = 5)
ggsave("../figs/niche_grid.pdf",         plot = p_occ / p_dens,            width = 4.5, height = 4.5)
ggsave("../figs/niche_breadth.pdf",      plot = p_hv,                      width = 3.5,  height = 2.5)
#ggsave("../figs/niche_heat_10.pdf",      plot = heat_plots[["p_heat_10"]], width = 8, height = 6)
ggsave("../figs/niche_heat_20.pdf",      plot = heat_plots[["p_heat_20"]], width = 4, height = 5)
ggsave("../figs/dens_Boxplot.pdf",         plot = p_dens_boxplot,                    width = 5, height = 3.5)



```
#write_csv(pcoa_coords,       "pcoa_coordinates.csv")
#write_csv(niche_summary,     "niche_count_results.csv")
#write_csv(niche_breadth,     "niche_breadth_results.csv")
#write_csv(grid_summary_all,  "grid_occupancy_results.csv")
write_csv(pcoa_clust %>% select(taxon_sample, sample_id, niche),
          "cluster_assignments.csv")

if (!use_fallback && exists("overlap_mat")) {
  as.data.frame(overlap_mat) %>%
    rownames_to_column("sample_id") %>%
    write_csv("hypervolume_pairwise_overlap.csv")
}

message("All outputs saved.")
```
